---
title: "Week 3: population in-out sim"
author: "Stephen R. Proulx"
date: "9/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

## A single simulation of immigration/emmigration 

Describe the model

```{r set_paramaters}
immi=0.15
emmi=0.2
tmax=5000
n0=40
```


Initialize a dataframe. Make a column for the time step, a column for the random number used to determine immigration, a column for the random number used to determine emmigration, and a column for the number of individuals. Just fill this last column with a single number, the initial populaiton size
```{r initialize_data}
popsize <- tibble(time=1:tmax , rand1=runif(tmax), rand2=runif(tmax), num=n0 )

```

```{r loop_sim}
for(i in 2:tmax){ 
  popsize$num[i] <- (immi>popsize$rand1[i])*1-(emmi>popsize$rand2[i])*1 + popsize$num[i-1]
  popsize$num[i]<- (popsize$num[i]<0)*1+(popsize$num[i]>0)*popsize$num[i]
}

```


```{r}
ggplot(data=popsize , aes(x=time,y=num)) + geom_point()
```

 
## Now do a bunch of simulations at once

Initialize a dataframe. Make a column for the time step, a column for the random number used to determine immigration, a column for the random number used to determine emmigration, and a column for the number of individuals. Just fill this last column with a single number, the initial populaiton size

how many simulations to do?
```{r}
k=20
```

```{r initialize_data}

popsize <- tibble(run=rep(1:k,each=tmax), time=rep(1:tmax,k) , rand1=runif(tmax*k), rand2=runif(tmax*k), num=n0 )

```

```{r loop_sim}
for(j in 1:k){
  for(i in 2:(tmax)){ 
    popsize$num[((j-1)*tmax+i)] <- (immi>popsize$rand1[((j-1)*tmax+i)])*1-(emmi>popsize$rand2[((j-1)*tmax+i)])*1 + popsize$num[((j-1)*tmax+i)-1]
    popsize$num[((j-1)*tmax+i)]<- (popsize$num[((j-1)*tmax+i)]<0)*1+(popsize$num[((j-1)*tmax+i)]>0)*popsize$num[((j-1)*tmax+i)]
  }
  }

```


```{r}
summarize_sims <- group_by(popsize, time) %>% summarize(
mean_n = mean(num),
lower_n = quantile(num, 0.10), upper_n = quantile(num, 0.90)
) %>% ungroup()
```

```{r}
ggplot(data=summarize_sims , aes(x=time,y=mean_n)) + 
  geom_point()+
  geom_ribbon(aes(ymin=lower_n,ymax=upper_n), alpha = 0.5, fill = "blue")
```

recode to get differences in pop size
```{r}
popsize<- popsize %>%
  group_by(run) %>%
  mutate(deltanum = num-lag(num)) %>%
  ungroup()

```

summarize by current popsize


```{r}
summarize_sims_by_n <- filter(popsize,time>1) %>%
  group_by(num) %>% summarize(
delta_n = mean(deltanum),
lower_n = quantile(deltanum, 0.10), upper_n = quantile(deltanum, 0.90)
) %>% ungroup()
```


```{r}
ggplot(data=summarize_sims_by_n , aes(x=num,y=delta_n)) + 
  geom_point()+
  geom_ribbon(aes(ymin=lower_n,ymax=upper_n), alpha = 0.5, fill = "blue")+
  ylim(c(-2,2))
```
